\chapter{Overview}
\section{Intended Audience}
This book presents a single semester course  on {\em fluid mechanics}\/ that is intended primarily for advanced undergraduate students majoring in physics. A thorough  understanding
of physics at the lower-division level, including a basic working
knowledge of the laws of mechanics, is assumed. It is
also taken for granted that students are
 familiar with the fundamentals of integral and differential
calculus,  complex analysis,  and ordinary differential equations. 
On the other hand, vector analysis plays such a central role in the study of fluid
mechanics that a brief, but fairly comprehensive, 
review of this subject area is provided in Appendix~\ref{vector}. 
Likewise, those aspects of cartesian tensor theory, orthogonal curvilinear coordinate systems, and
the calculus of variations, that are required in the study of fluid mechanics are outlined in Appendices~\ref{appb}, \ref{ccurv}, and \ref{cvar}, respectively.

\section{Outline of Course}
{\em Fluid mechanics}\/ is an area of study that has applications to a wide range of different disciplines, including civil
engineering, chemical engineering, atmospheric and oceanic physics, geophysics, space physics, 
astrophysics, biophysics, and medical physics. This course concentrates on that limited, but nevertheless very large, subset of these applications which is amenable to conventional mathematical analysis. 

\section{Major Sources}
The material appearing in Appendix~\ref{vector} is largely based on the author's recollections
of a vector analysis course given by Dr.~Stephen Gull at the University
of Cambridge. Major sources for the material appearing in other chapters and appendices 
include:
\begin{description}
\item [\em Statics, Including Hydrostatics and the Elements of the Theory of Elasticity] H.~Lamb, 3rd Edition (Cambridge University Press, Cambridge UK, 1928).
\item [\em Hydrodynamics] H.~Lamb, 6th Edition (Dover, New York NY, 1945).
\item[\em Theoretical Aerodynamics] L.M.~Milne-Thomson, 4th Edition, Revised and enlarged (Dover, New York NY, 1958).
\item [\em Ellipsoidal Figures of Equilibrium] S.~Chandrasekhar (Yale University Press, New Haven CT, 1969).
\item[\em Boundary Layer Theory] H.~Schlichting 7th Edition (McGraw-Hill, New York NY, 1970).
\item [\em Mathematical Methods for the Physical Sciences] K.F.~Riley (Cambridge University Press, Cambridge UK, 1974).
\item [\em Fluid Mechanics] L.D.~Landau, and E.M.~Lifshitz, 2nd Edition, (Butterworth-Heinemann, Oxford UK, 1987).
\item [\em Physical Fluid Dynamics] D.J.~Tritton, 2nd Edition (Oxford University Press, Oxford UK, 1988).
\item [\em Fluid Dynamics for Physicists] T.E.~Faber, 1st Edition (Cambridge University Press, Cambridge UK, 1995).
\item [\em Schaum's Outline of Fluid Dynamics] W.~Hughes, and J.~Brighton, 3rd Edition (McGraw-Hill, New York NY, 1999).
\item [\em An Introduction to Fluid Dynamics] G.K.~Batchelor (Cambridge University Press, Cambridge UK, 2000).
\item [\em Theoretical Hydrodynamics] L.M.~Milne-Thomson, 5th Edition (Dover, New York NY, 2011).
\end{description}

